Computational methods in landscape ecology

### 2024-09-20 Manuscript.qmd
Adding comments AdF

author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affil-id: 1
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affil-id: 2,3
  - name: Craig E. Simpkins
    orcid: 0000-0003-3212-1379
    email: simpkinscraig063@gmail.com
    affil-id: 4
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affil-id: 1
  - name: Gemma Gerber
    orcid: 0000-0002-8722-3113
    email: gerber@iiasa.ac.at
    affil-id: 1
  - name: Alida de Flamingh
    orcid: 0000-0003-1223-6654
    email: adeflamingh@gmail.com
    affil-id: 5
  - name: Martí Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affil-id: 6

affiliations:
  - id: 1
    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - id: 2
    name: Adam Mickiewicz University, Poznań, Poland
  - id: 3
    name: University of Münster, Münster, Germany
  - id: 4
    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
  - id: 5
    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
  - id: 6
    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland

@@

## References {.unnumbered}

>>>>

author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affil-id: 1
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affil-id: 2,3
  - name: Craig E. Simpkins
    orcid: 0000-0003-3212-1379
    email: simpkinscraig063@gmail.com
    affil-id: 4
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affil-id: 1
  - name: Alida de Flamingh
    orcid: 0000-0003-1223-6654
    email: adeflamingh@gmail.com
    affil-id: 5
  - name: Gemma Gerber
    orcid: 0000-0002-8722-3113
    email: gerber@iiasa.ac.at
    affil-id: 1
  - name: Martí Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affil-id: 6

affiliations:
  - id: 1
    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - id: 2
    name: Adam Mickiewicz University, Poznań, Poland
  - id: 3
    name: University of Münster, Münster, Germany
  - id: 4
    name: Auckland Council, Environmental Services, Auckland, New Zealand
  - id: 5
    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
  - id: 6
    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland

@@

# References {.unnumbered}

### 2024-09-20 Sections/01-introduction.qmd
Adding comments AdF

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
![Overview of landscape ecology topics for which computational developments are discussed.](Fig-1.png)

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introduction to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

![Overview of landscape ecology topics for which computational developments are discussed.](Figures/Fig-1.png){width=85%}

### 2024-09-20 Sections/02-spatial-patterns.qmd
Adding comments AdF

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, such as correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency.
These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
<!-- MH: I moved sections around quite a bit unfortunately -->

@@

Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics.
They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
However, studies show that insights do not only depend on the formulation of the selected entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].

@@

Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
@Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the landscape complexity.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC classes are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

Surface metrics are based on the gradient surface model using gridded raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
The gradient surface model can increase the resemblance of the data and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].

Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
Yet, nowadays, vector data is increasingly available and has been widely used in other fields, such as geography or urban studies.

Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].

However, also vector-based metrics have their limitations.
The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.

@@

Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures).
Spatial patterns can be compared between different areas, or for the same area at two different moments in time.

Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].

>>>>

Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are uncomplicated to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, for example correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
These components of landscape patterns are in line with revent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency, and that these components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.

@@

Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics.
They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
However, studies show that insights gained from entropy measures depend on the formulation of the selected entropy, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].

@@

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
@Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify the landscape complexity.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases where the dissimilarities between LULC classes are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

Surface metrics are based on the gradient surface model using gridded raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].

Yet, software to calculate these metrics are still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].

@@

Many landscape ecology studies are based on the raster data model.
However, the choice of the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
However, nowadays, vector data has become increased in availability and has been widely used in other fields, such as geography or urban studies.

Hundreds of metrics have already been developed for raster data, and the most straightforward approach may therefore be to reimplement these same metrics for vector data.
Although this is possible for most of the metrics, there are a few exceptions where metrics are specifically related to the raster data model [@Yao2022VecLIFramework].
For example, in urban planning, vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].

However, also vector-based metrics also have limitations.
The most important limitations relate to computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often especially problematic for data from different sources.
Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.

@@

Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures).
Spatial patterns can be compared between different areas, or for the same area at two different moments in time.

Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
This allows for the identification of areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].
